quollr: An R Package for Visualizing 2-D Models from Non-linear Dimension Reductions in High Dimensional Space

### 2025-09-17 paper-quollr.Rmd
Updated glance and augment function implementations

<!-- If a user would like to perform steps of the algorithm themselves, additional user input will be needed for the function that perform each step. For example, if the user wishes to use already binning data, the `extract_hexbin_centroids()` function can be used directly. -->

@@

The `extract_hexbin_centroids()` function combines hexagonal bin coordinates, raw and standardized counts within each hexagons.

@@

```{r, echo=TRUE}
df_bin_centroids <- extract_hexbin_centroids(
  centroids_data = all_centroids_df, 
  counts_data = hb_obj$std_cts
  )

@@

```{r, echo=TRUE}
find_low_dens_hex(
  model_2d = df_bin_centroids, 
  b1 = 15, 
  benchmark_mean_dens = 0.05
)

@@

```{r, echo=TRUE}
glance(
  highd_data = scurve, 
  model_2d = df_bin_centroids, 
  model_highd = df_bin
  )
```

@@

```{r, echo=TRUE}
model_error <- augment(
  highd_data = scurve, 
  model_2d = df_bin_centroids, 
  model_highd = df_bin
  )

>>>>

<!-- If a user would like to perform steps of the algorithm themselves, additional user input will be needed for the function that perform each step. For example, if the user wishes to use already binning data, the `merge_hexbin_centroids()` function can be used directly. -->

@@

The `merge_hexbin_centroids()` function combines hexagonal bin coordinates, raw and standardized counts within each hexagons.

@@

```{r, echo=TRUE}
df_bin_centroids <- merge_hexbin_centroids(
  centroids_data = all_centroids_df, 
  counts_data = hb_obj$std_cts
  )

@@

```{r, echo=TRUE}
find_low_dens_hex(
  model_2d = df_bin_centroids, 
  b1 = 15, 
  md_thresh = 0.05
)

@@

```{r, echo=TRUE}
glance(
  model_object = scurve_model_obj,
  highd_data = scurve
  )
```

@@

```{r, echo=TRUE}
model_error <- augment(
  model_object = scurve_model_obj,
  highd_data = scurve
  )

### 2025-09-17 paper-quollr.tex
Updated glance and augment function implementations

\begin{figure}[H]
\includegraphics[width=1\linewidth]{paper-quollr_files/figure-latex/overview-1} \caption{Wireframe model representation of the NLDR layout, lifted and displayed in high-dimensional space. The left panel shows the NLDR layout with a triangular mesh overlay, forming the wireframe structure. This mesh can be lifted into higher dimensions and projected to examine how the geometric structure of the data is preserved. Panels (a1–a4) display different $2 ext{-}D$ projections of the lifted wireframe, where the underlying curved sheet structure of the data is more clearly visible. The triangulated mesh highlights how local neighborhoods in the layout correspond to relationships in the high-dimensional space, enabling diagnostics of distortion and preservation across dimensions.}\label{fig:overview}
\end{figure}

@@

The remaining variables \(X_4\), \(X_5\), \(X_6\), \(X_7\) are all uniform error, with small variance. \texttt{scurve\_umap} is the UMAP \(2\text{-}D\) embedding for \texttt{scurve} data with \texttt{n\_neighbors}\(=46\) and \texttt{min\_dist}\(=0.9\). Each data set contains a unique ID column that maps the \(7\text{-}D\) point in \texttt{scurve} to the corresponding \(2\text{-}D\) \texttt{scurve\_umap}.

@@

This function requires several parameters: the high-dimensional data (\texttt{highd\_data}), the emdedding data (\texttt{nldr\_data}), the number of bins along the x-axis (\texttt{b1}), the buffer amount as a proportion of data (\texttt{q}), and benchmark value to extract high density hexagons (\texttt{benchmark\_highdens}). The function returns an object that includes the scaled NLDR object (\texttt{nldr\_obj}), the hexagonal object (\texttt{hb\_obj}), the fitted model in both \(2\text{-}D\) (\texttt{model\_2d}), and \(p\text{-}D\) (\texttt{model\_highd}), and triangular mesh (\texttt{trimesh\_data}).

\begin{verbatim}
fit_highd_model(
  highd_data = scurve, 
  nldr_data = scurve_umap, 
  b1 = 15, 
  q = 0.1, 
  benchmark_highdens = 1)
\end{verbatim}

@@

The algorithm starts by scaling the NLDR data to to the range \([0, 1] \times [0, x_2]\), where \(x_2\) is the maximum value of the second embedding, using the \texttt{gen\_scaled\_data()} function. The output includes the scaled NLDR data (\texttt{scaled\_nldr}) along with the original limits of the embeddings (\texttt{lim1}, \texttt{lim2}).

@@

The configurations of a hexagonal grid are determined by the number of bins and the bin width in each direction. The function \texttt{calc\_bins\_y()} is used for this purpose. This function accepts (1) an object containing three elements: the first is the scaled NLDR data with two columns, and the second and third are the limits of the original NLDR data; (2) the number of bins along the x-axis (\texttt{b1}), and (3) the buffer amount as a proportion (\texttt{q}).

\begin{verbatim}
bin_configs <- calc_bins_y(
  nldr_obj = scurve_umap_obj, 
  b1 = 15, 
  q = 0.1)

@@

Points are allocated to bins based on the nearest centroid of the hexagonal bins. The hexagonal binning algorithm can be executed using the \texttt{hex\_binning()} function, or its components can be run separately for added flexibility. The parameters used within \texttt{hex\_binning()} include an object containing scaled NLDR data in the first and second columns, along with numeric vectors that represent the limits of the original NLDR data (\texttt{nldr\_obj}), the number of bins along the x-axis (\texttt{b1}), and the buffer amount as a proportion of the data (\texttt{q}). The output is an object of the \texttt{hex\_bin\_obj} class, which contains the bin widths in each direction (\texttt{a1}, \texttt{a2}), the number of bins in each direction (\texttt{bins}), the coordinates of the hexagonal grid starting point (\texttt{start\_point}), the details of bin centroids (\texttt{centroids}), the coordinates of bins (\texttt{hex\_poly}), NLDR components with their corresponding hexagon IDs (\texttt{data\_hb\_id}), hex bins with their corresponding standardized counts (\texttt{std\_cts}), the total number of bins (\texttt{tot\_bins}), the number of non-empty bins (\texttt{non\_bins}), and the points within each hexagon (\texttt{pts\_bins}).

\begin{verbatim}
hb_obj <- hex_binning(
  nldr_obj = scurve_umap_obj, 
  b1 = 15, 
  q = 0.1)
\end{verbatim}

{\centering \includegraphics[width=0.5\linewidth]{paper-quollr_files/figure-latex/hex-param-1}

\caption{The components of the hexagon grid illustrating notation.}\label{fig:hex-param}
\end{figure}

If the hexagonal binning process is run separately, it involves several steps: (i) generating all possible centroids in a hexagonal grid, (ii) creating the coordinates of the hexagons, (iii) assigning data points to their respective hexagons, (iv) computing the standardized number of points within each hexagon, and (v) mapping the points to their corresponding hexagonal bins.

@@

\begin{verbatim}
all_centroids_df <- gen_centroids(
  nldr_obj = scurve_umap_obj, 
  b1 = 15, 
  q = 0.1
  )

@@

For each centroid, six vertices are computed and assigned a polygon ID (\texttt{h}) corresponding to the centroid's \texttt{h}. These points are aggregated into a tibble that includes the polygon ID (\texttt{h}) and the respective x (\texttt{x}) and y (\texttt{y}) coordinates for all hexagon corners.

@@

\begin{verbatim}
umap_hex_id <- assign_data(
  nldr_obj = scurve_umap_obj, 
  centroids_data = all_centroids_df
  )

@@

The function begins by grouping the data by hexagon ID (\texttt{h}) and counting the number of NLDR points falling within each bin. These raw counts are stored as \texttt{n\_h}. To enable comparisons across bins with varying densities, the function then standardizes these counts by dividing each bin's count by the maximum count across all bins. This yields a normalized metric, \texttt{w\_h}, ranging from 0 to 1.

@@

The \texttt{find\_pts()} function extracts the list of data point identifiers (\texttt{ID}) assigned to each hexagon in the NLDR space.

@@

\begin{verbatim}
pts_df <- find_pts(
  scaled_nldr_hexid = umap_hex_id
  )

@@

The \texttt{extract\_hexbin\_centroids()} function combines hexagonal bin coordinates, raw and standardized counts within each hexagons.

@@

\begin{verbatim}
df_bin_centroids <- extract_hexbin_centroids(
  centroids_data = all_centroids_df, 
  counts_data = hb_obj$std_cts
  )

@@

The \texttt{tri\_bin\_centroids()} function generates a triangulation object from the x and y coordinates of hexagon centroids using the \texttt{tripack::tri.mesh()} function. This triangulation forms the structural basis for identifying adjacent bins.

@@

@@

\begin{verbatim}
find_low_dens_hex(
  model_2d = df_bin_centroids, 
  b1 = 15, 
  benchmark_mean_dens = 0.05
)
\end{verbatim}

@@

For simplicity, we remove low-density hexagons using a threshold of \(10\).

@@

@@

\begin{verbatim}
glance(
  highd_data = scurve, 
  model_2d = df_bin_centroids, 
  model_highd = df_bin
  )
\end{verbatim}

\begin{verbatim}
> # A tibble: 1 x 2
>   Error  RMSE
>   <dbl> <dbl>
> 1  265. 0.157
\end{verbatim}

@@

\begin{verbatim}
model_error <- augment(
  highd_data = scurve, 
  model_2d = df_bin_centroids, 
  model_highd = df_bin
  )

>>>>

\begin{figure}[H]
\includegraphics[width=1\linewidth]{paper-quollr_files/figure-latex/overview-1} \caption{Wireframe model representation of the NLDR layout, lifted and displayed in high-dimensional space. The left panel shows the NLDR layout with a triangular mesh overlay, forming the wireframe structure. This mesh can be lifted into higher dimensions and projected to examine how the geometric structure of the data is preserved. Panels (a1–a4) display different $2\text{-}D$ projections of the lifted wireframe, where the underlying curved sheet structure of the data is more clearly visible. The triangulated mesh highlights how local neighborhoods in the layout correspond to relationships in the high-dimensional space, enabling diagnostics of distortion and preservation across dimensions.}\label{fig:overview}
\end{figure}

@@

The remaining variables \(X_4\), \(X_5\), \(X_6\), and \(X_7\) are included as low-magnitude noise dimensions, generated from uniform distributions with narrow ranges around zero: \(X_4, X_5 \sim U(-0.02, 0.02)\), \(X_6 \sim U(-0.1, 0.1)\), and \(X_7 \sim U(-0.01, 0.01)\). These small variances ensure that the added dimensions contribute noise without obscuring the main geometric structure. \texttt{scurve\_umap} is the UMAP \(2\text{-}D\) embedding for \texttt{scurve} data with \texttt{n\_neighbors}\(=46\) and \texttt{min\_dist}\(=0.9\). Each data set contains a unique ID column that maps the \(7\text{-}D\) point in \texttt{scurve} to the corresponding \(2\text{-}D\) \texttt{scurve\_umap}.

@@

This function requires several parameters: the high-dimensional data (\texttt{highd\_data}), the emdedding data (\texttt{nldr\_data}), the number of bins along the x-axis (\texttt{b1}), the buffer amount as a proportion of data (\texttt{q}), and benchmark value to extract high density hexagons (\texttt{hd\_thresh}). The function returns an object (\texttt{nldr\_scaled\_obj}) containing three elements: the first is the scaled NLDR data (\texttt{scaled\_nldr}), and the second and third are the limits of the original NLDR data (\texttt{lim1} and \texttt{lim2}); the hexagonal object (\texttt{hb\_obj}), the fitted model in both \(2\text{-}D\) (\texttt{model\_2d}), and \(p\text{-}D\) (\texttt{model\_highd}), and triangular mesh (\texttt{trimesh\_data}).

\begin{verbatim}
fit_highd_model(
  highd_data = scurve, 
  nldr_data = scurve_umap, 
  b1 = 15, 
  q = 0.1, 
  hd_thresh = 1)
\end{verbatim}

@@

The algorithm starts by scaling the NLDR data to to the range \([0, 1] \times [0, y_{2,max}]\), where \(y_{2,max} = r_2/r_1\) is the ratio of ranges of embedding components. The output includes the scaled NLDR data (\texttt{scaled\_nldr}) along with the original limits of the embeddings (\texttt{lim1}, \texttt{lim2}).

@@

The configurations of a hexagonal grid are determined by the number of bins and the bin width in each direction. The function \texttt{calc\_bins\_y()} is used for this purpose. This function accepts (1) an object (\texttt{nldr\_scaled\_obj}) containing three elements: the first is the scaled NLDR data (\texttt{scaled\_nldr}), and the second and third are the limits of the original NLDR data (\texttt{lim1} and \texttt{lim2}); (2) the number of bins along the x-axis (\texttt{b1}), and (3) the buffer amount as a proportion (\texttt{q}). The buffer ensures that the grid fully covers the data space by extending one hexagon width (\(a_1\)) and height (\(a_2\)) beyond the observed data in all directions. By default, \(q = 0.1\), but it must be set to a value smaller than the minimum data value to avoid exceeding the data range.

\begin{verbatim}
bin_configs <- calc_bins_y(
  nldr_scaled_obj = scurve_umap_obj, 
  b1 = 15, 
  q = 0.1)

@@

Points are allocated to bins based on the nearest centroid of the hexagonal bins. The hexagonal binning algorithm can be executed using the \texttt{hex\_binning()} function, or its individual components can be run separately for added flexibility. While running the process step by step would involve generating centroids, constructing hexagon coordinates, assigning points to bins, standardizing counts, and mapping the data back to hexagons, the \texttt{hex\_binning()} function automates this entire workflow. The parameters used within \texttt{hex\_binning()} include an object (\texttt{nldr\_scaled\_obj}) containing three elements: the first is the scaled NLDR data (\texttt{scaled\_nldr}), and the second and third are the limits of the original NLDR data (\texttt{lim1} and \texttt{lim2}); the number of bins along the x-axis (\texttt{b1}), and the buffer amount as a proportion of the data (\texttt{q}). The output is an object of the \texttt{hex\_bin\_obj} class, which contains the bin widths in each direction (\texttt{a1}, \texttt{a2}), the number of bins in each direction (\texttt{bins}), the coordinates of the hexagonal grid starting point (\texttt{start\_point}), the details of bin centroids (\texttt{centroids}), the coordinates of bins (\texttt{hex\_poly}), NLDR components with their corresponding hexagon IDs (\texttt{data\_hb\_id}), hex bins with their corresponding standardized counts (\texttt{std\_cts}), the total number of bins (\texttt{tot\_bins}), the number of non-empty bins (\texttt{non\_bins}), and the points within each hexagon (\texttt{pts\_bins}).

\begin{verbatim}
hb_obj <- hex_binning(
  nldr_scaled_obj = scurve_umap_obj, 
  b1 = 15, 
  q = 0.1)
\end{verbatim}

@@

\begin{verbatim}
all_centroids_df <- gen_centroids(
  nldr_scaled_obj = scurve_umap_obj, 
  b1 = 15, 
  q = 0.1
  )

@@

For each centroid, six vertices are computed and assigned a polygon ID as the centroid. These vertices are then combined into a tibble that records the polygon ID (\texttt{h}) and the respective x (\texttt{x}) and y (\texttt{y}) coordinates for all hexagon corners.

@@

\begin{verbatim}
umap_hex_id <- assign_data(
  nldr_scaled_obj = scurve_umap_obj, 
  centroids_data = all_centroids_df
  )

@@

The function begins by grouping the data by hexagon ID (\texttt{h}) and counting the number of NLDR points falling within each bin. These raw counts are stored as \texttt{n\_h}. To enable comparisons across bins with varying densities, the function then standardizes these counts by dividing each bin's count by the maximum count across all bins. This yields a standardized bin counts, \texttt{w\_h}, ranging from 0 to 1.

@@

The \texttt{group\_hex\_pts()} function extracts the list of data point identifiers (\texttt{ID}) assigned to each hexagon in the NLDR space.

@@

\begin{verbatim}
pts_df <- group_hex_pts(
  scaled_nldr_hexid = umap_hex_id
  )

@@

The \texttt{merge\_hexbin\_centroids()} function combines hexagonal bin coordinates, raw and standardized counts within each hexagons.

@@

\begin{verbatim}
df_bin_centroids <- merge_hexbin_centroids(
  centroids_data = all_centroids_df, 
  counts_data = hb_obj$std_cts
  )

@@

The \texttt{tri\_bin\_centroids()} function generates a triangulation object from the x and y coordinates of hexagon centroids using the \texttt{tripack::tri.mesh()} function (\citet{albrecht2025}). This triangulation forms the structural basis for identifying adjacent bins.

@@

\begin{verbatim}
> # A tibble: 5 x 10
>    from    to  x_from  y_from    x_to    y_to from_count to_count from_reindexed
>   <int> <int>   <dbl>   <dbl>   <dbl>   <dbl>      <dbl>    <dbl>          <int>
> 1     1     2 -0.1    -0.116  -0.0166 -0.116           0        0              1
> 2    16    17 -0.0583 -0.0434  0.0251 -0.0434          0        0             16
> 3    16    32 -0.0583 -0.0434 -0.0166  0.0288          0        0             16
> 4     3    17  0.0668 -0.116   0.0251 -0.0434          0        0              3
> 5    32    33 -0.0166  0.0288  0.0668  0.0288          0        0             32
> # i 1 more variable: to_reindexed <int>
\end{verbatim}

@@

\begin{verbatim}
find_low_dens_hex(
  model_2d = df_bin_centroids, 
  b1 = 15, 
  md_thresh = 0.05
)
\end{verbatim}

@@

For simplicity, we remove low-density hexagons using a threshold of \(1\).

@@

It is worth noting that while \texttt{predict\_emb()} provides a general approach that works across methods, some NLDR techniques have their own built-in prediction mechanisms. For example, UMAP (\citet{tomasz2023}) supports direct prediction of embeddings for new data once a model is fitted.

@@

\begin{verbatim}
glance(
  model_object = scurve_model_obj,
  highd_data = scurve
  )
\end{verbatim}

\begin{verbatim}
> # A tibble: 1 x 2
>   Error  RMSE
>   <dbl> <dbl>
> 1  273. 0.163
\end{verbatim}

@@

\begin{verbatim}
model_error <- augment(
  model_object = scurve_model_obj,
  highd_data = scurve
  )
